\newpage\nuluj
\pagestyle{sever}

\logo

\nadpis{Protecting Human Rights Through Fundamental\\Principles of Administrative Procedures\\in Eastern Europe}

\autor{Tina Sever\pozn{University of Ljubljana, Faculty of Administration, Gosarjeva ul.\;5, 1000 Ljubljana, Slovenia. E-mail:\;tina.se\-ver@fu.uni-lj.si.}, Iztok Rakar\pozn{University of Ljubljana, Faculty of Administration, Gosarjeva ul.\;5, 1000 Ljubljana, Slovenia. E-mail: iztok.ra\-kar@fu.uni-lj.si.}, Polonca Kovaè\pozn{University of Ljubljana, Faculty of Administration, Gosarjeva ul.\;5, 1000 Ljubljana, Slovenia. E-mail: polo\-na.kovac@fu.uni-lj.si.}}
\linka{3ex}

\cast{Abstract}
Administrative procedures (APs) are tools to protect fundamental human rights in state-citizen relations. As the modernization of public administration regulation is undergoing a~transformation in the direction of reducing detailed rules on APs and, by the same token, emphasizing fundamental or general principles, research on the development and the state of the art of administrative principles in the general administrative procedure acts (APAs) of selected Eastern Europe countries with a~common heritage of Austrian law dating back to 1925 (Slovenia, Croatia, Macedonia and the Czech Republic) was carried out. The normative-comparative analysis reveals differences in approaches to and the pace of APAs reform and content; some countries are taking a~more radical approach, mainly by following good governance dimensions. Convergence based on Council of Europe and EU initiatives is also evident. Classical guarantees against the misuse of power (principles of legality, equality, proportionality, rights of defense, etc.) are therefore crucial. The most progress seems to have been made by Croatia and the Czech Republic; by focusing on partnerships in administrative-legal relations in the sense of good administration, these two countries have, among other things, set a~trend for other countries to follow.\\

\cast{Keywords}
Administrative Procedures, Eastern Europe, Good Governance, Human Rights, Principles, Rule of Law
\linka{4ex}

\clearpage\newpage

\podnadpis{Introduction}
\oddil{Human Rights Protection in Administrative Relations}
The historical conflict between the interests of the state and the citizen has led to an ever-evolving process of human rights protection in law and practice. In the time of absolute monarchies, the authority executed its powers \fntit{ex imperio} with no acknowledgment of individual rights, as the latter were unilaterally subordinate to the state. The existing regulations were intended to support the state apparatus and the execution of its powers. These trends changed with the rise of the bourgeoisie, wherein the French revolution (1789--1799) was a~milestone. This was a~very important event for the development of administrative law and the origins of individual rights. One result of the French Revolution was the codification of human rights in the Declaration of the Rights of Man and of the Citizen. The Declaration codified basic principles that established new approaches to the state-citizen relation as a~fundamental administrative relation, for example: \fntit{``The Law has the right to forbid only those actions that are injurious to society.''}\pozn{And: \fntit{``Nothing that is not forbidden by Law may be hindered, and no one may be compelled to do what the Law does not ordain''} (Article 5). Furthermore, Article 12: \fntit{``To guarantee the Rights of Man and of the Citizen a~public force is necessary; this force is therefore established for the benefit of all, and not for the particular use of those to whom it is entrusted.''} English version source available at Britannica Academic (2014).} The precondition for the rule of law and the freedom of the individual and his/her human rights and fundamental freedoms is the separation of powers; this was already recognized by Montesquieu (1689 to 1755).\pozn{Ker¹evan (2008).}
%%%%%%%% tady jsme zmenili -- 

As a~counterweight to absolutism and a~police state (\fntit{Polizeistaat}), a~conception of the rule of law (\fntit{Rechtsstaat}) rooted in the Age of Enlightenment was also outlined by nineteenth-century German writers. Initially (up to the twentieth century), the idea concentrated on legal formalism and was linked to a~bureaucratic apparatus which guaranteed the functioning of the system. But with World War II and the Nazi regime, which abused the law by adopting tyrannical rules, the idea of \fntit{Rechtsstaat} had to be significantly changed to include not only formal legality (which is important for procedural safeguards) but also the constitutionally recognized protection of human rights under the supervision of the constitutional court and furthermore good (public) governance.\pozn{Ziller (2009, cop. 2007), cf. Pollitt and Bouckaert (2011).} This doctrine is of major importance within the research presented in this paper since most of the states in the area of Eastern Europe were once under German and Austrian influence in the administrative field (Slovenia, the Czech Republic, and Croatia). The German understanding of the rule of law is more closely related to the separation of powers and to administration being bound by law than the American one, for example\pozn{Cf. Künnecke (2007), Pavènik (2009).}. \fntit{Rechtsstaat} implies the elimination of arbitrary authority and is the ideal of the democratic state. Yet in order to implement the concept of the rule of law, a~legalistic approach (e.g. the constitutional provision on the independence of judges) does not suffice and it is necessary to examine and act appropriately in the implementation of legal norms in practice. In this framework the administrative procedures in most of Eastern Europe are understood as concrete \mbox{decision-making} in individual matters only (so-called \fntit{adjudication} or \fntit{Verwaltungsverfahren}, as opposed to general policies decision-making)\pozn{Cf. Rusch (2009), Galligan, Langan, and Constance (1998).}. The core of administrative constitutional and supranational law is due process or fair trial. Administrative procedural law was historically developed in the individual states primarily to protect the fundamental human rights of parties in relation to the (mis)use of power by administrative authorities (as \fntit{constitutional law in action}\pozn{See Künnecke (2007).}).

The authors have put forth the hypothesis that administrative procedural law principles are a~tool for making operational the basic constitutional safeguards for the protection of individual human rights from the abuse of power when deciding on administrative matters. Fundamental principles, such as the principles of legality, proportionality and equality, are the product of societal development over time and are inextricably linked to the promotion and protection of human rights. Principles therefore represent the fundamental values valid for a~certain space and time in society. The mutual influence of the legal orders of different national systems is evident throughout history and becomes even more evident after World War II, when international organizations to promote the protection of human rights, such as the United Nations and the Council of Europe, were established. Fundamental principles are the core interpretative instrument of legal rules and as such they are a~source of and guideline for human rights protection. The corpus of fundamental principles is not a~status quo, but an ever-evolving process that is also linked to technological progress. In the last century, the  technological progress that made the most visible impact on society has been the Internet, which enabled many different kinds of e-communication and necessitated the regulation thereof and with it the promotion of principles of participation, transparency, efficiency, removing administrative burdens, i.e. principles of good administration. With the overall globalization process taking part in different segments of society, including law, and on the basis of the similar historical background of the selected Eastern European states, we defend the idea that a~convergence of administrative procedure regulation is taking place and promoting those same values which are enshrined in fundamental principles. 

\oddil{Research Question and Research Methodology }
Administrative procedure (AP) is one of the most important instruments through which parties can receive substantive rights in relation to the state. One of the goals of its codification is to ensure procedural standards for the parties and protect their rights (procedural as well as substantive). Fundamental principles as a~guideline on the interpretation of legal rules are important and common to all legal branches (principles of legality such as proportionality and equality are already commonly recognized by constitutions). Furthermore, administrative law, with its sub and supra position of party and authority, demands certain ``special'' (or with certain modifications for the executive branch) principles, such as substantive truth, the protection of public interest, independence, etc.

The central research question is whether fundamental principles of APs are important for the protection of human rights. Therefore we studied the following research sub-questions: firstly, the setting up of a~model of fundamental principles of good administration and good governance as deriving from different international documents, presenting basic pool of values to be followed by APs' regulation. Secondly, the identification of existing fundamental principles in APs' codification in selected states, and thirdly, the assessment of their relevance for human rights protection. The authors investigated the level of human rights protection in APs regulation of Croatia, the Czech Republic, Slovenia and the Former Yugoslav Republic of Macedonia (hereinafter referred to as Macedonia), focusing on variations and similarities/differences among the states and pace of progress achieved after the downfall of communist regimes in the 1990s with the influence of joining (or being in the process of joining) the supranational authority of EU, and the promoting of modern approaches. 

The sample of countries was chosen based on the facts as follows. The selected states are similar -- compared to the majority of Western European states -- in terms of size (they are relatively small), political culture, politicization, macro-economic criteria, etc.  They all belong to the so-called continental Europe state tradition, in which state dominates society and the understanding of public administration has its roots in public law -- this understanding emphasizes the role of the state and parliamentary legislation in defining policy over and above society.\pozn{Loughlin, Hendriks and Lindström (eds.) (2011).} One of the reasons for these similarities is a~common historical background. Croatia, the Czech Republic and Slovenia were all part of the Austro-Hungarian empire, which bequeathed a~certain level of common legacy to all three states. Furthermore, in 1925, Austria was regarded as the first country to successfully codify AP (after a~Spanish attempt to codify it in 1889). Selected countries in its geographical neighborhood and with a~previous common legacy, took over the Austrian model of APs regulation (Czechoslovakia in 1928 and Yugoslavia in 1930). There was a~consequent Austrian influence also on Macedonia, whose territory became part of the Kingdom of Yugoslavia in 1929. Later, the sample of countries was identified by all four countries being influenced by communist government after the World War II, with Czechoslovakia having been under Soviet influence, while the other three were under the common Yugoslavian socialist system. Since the 1990s, these countries have been aiming at the modernization of their regulation and development of democracy, being also influenced by efforts to join the EU and its demands for the modernization of legal order, the signing of the European Convention on Human Rights, etc. All four countries are therefore ``New Democracies'' and part of a~convergence wave. However, the research question is to what extent the abovementioned circumstances have influenced the regulation of fundamental principles of APs as human rights promoters and identifying commonalities and divergences among the compared states. 

In order to identify the latter, an analysis of general laws on administrative procedures in four selected states was carried out, with the research focusing on the introductory provisions that regulate fundamental principles. They are as follows:
\zacseznam
\item \fntit{Zakon o~splo¹nem upravnem postopku}, Slovene APA (Official Gazette No. 80/99 and amendments);

\clearpage\newpage

\item \fntit{Zakon o~op¹tem upravnom postupku}, Macedonian APA (Official Gazette No. 38/05 and amendments);
\item \fntit{Zákon è. 500/2004 Sb., správní øád} (Act. No. 500/2004 Coll., Administrative Procedure Code);
\item \fntit{Zakon o~opæem upravnom postupku}, Croatian APA (Official Gazette No. 47/09).
\konseznam
For the purposes of the performed research, different methodologies were applied. In the introductory chapters, we used descriptive methodology giving an overview of the idea of the rule of law and the origins of human rights protection, with a~short historical introduction and the notion of the Germanic and Anglo-Saxon system, placing administrative procedural law in the role of human rights guardian in relation to administrative authorities. Using the historical method, we analyzed the development of generations of human rights over time with defining types of human rights relevant for each generation. Subsequently, a~comprehensive overview of good governance and good administration concepts is given by means of a~normative-analytical as well as historical method. Using a~systematic approach, the authors analyzed major international documents over time and, by means of methods of deduction and synthesis, drew up a~set of principles as a~model for the setting up of fundamental principles of APs. From the set of most common principles distinctive for each concept, classical constitutional safeguards and managerial and modern principles were identified by typological method. Based on legal theory, the authors explained the basic characteristics of principles and their relation to rules. In the second part of the article, the authors used a~historical and normative-analytical as well as comparative method to analyze AP regulation in selected states, using a~macro approach to first define AP codifications in general and their extent of use. Furthermore, the fundamental principles of APs were analyzed on a~micro level. With the use of the deductive method, normative analysis and analysis of content fundamental principles were identified of a~classical and managerial/modern nature, as deriving from APs of selected states. Their content was compared and interpreted, using methods of analogy and deduction as well as typology. Finally, using the axiological-deontological method, principles as promoters of human rights protection in APs were defined. In the conclusion a~synthesis of the findings was performed. 

\podnadpis{Protection of Human Rights and Public Interest\\in Administrative Procedures}
An authority in a~given time and space defines the public interest to be followed in a~certain state. As a~precondition for a~democratic state, public interest should be justified on the basis of commonly recognized values and the needs of the society in question. Protection of human rights should be a~component of this interest. In theory, different generations of human rights can be distinguished, in line with changes in the development of society over time. The protection of citizen and political rights (such as freedom of movement, personal dignity, the inviolability of human life) is typical of the first generation.\pozn{The main relevant international document is the International Covenant on Civil and Political Rights (1966, UN General Assembly). } These so-called \fntit{rights of negative status} demand \fntit{negative status duties} of the state, meaning that the state should suspend its intervention in the field of the beneficiaries of these rights. Nevertheless, a~certain degree of state action (and, to this extent, also positive duties) is expected for effectively ensuring these rights. The second generation of human rights is characterized by social and economic rights.\pozn{The main relevant international document relevant is the International Covenant on Economic, Social and Cultural Rights (1966).} These are rights of positive status, meaning that active action is expected of the state. The state should establish actual and legal conditions for individuals to be able to enjoy and execute these rights. Theory further delineates a~third generation of collective rights deriving from the collective right to self-determination, and a~fourth generation, which was the result of the development of post-industrial features (for example, the right to a~healthy living environment). The last two groups are often acknowledged in administrative procedures.\pozn{Lampe (2010).} 

Administrative procedures are a~means for providing principles of good administration and as such they establish an important part of the quality of public administration.\pozn{According to Rusch (2011), the current principles of good administration are reliability and predictability (respecting the rule of law, also by discretion decision-making; also decision-making in a~reasonable timeframe); openness and transparency (use of simple, clear language; encouragement of participation of everyone affected by a~decision; right to be heard); accountability (supervision by other branches of authority; possibility of legal remedies; giving reasons for a~decision; access to files, etc.); judicial control system and effectiveness and efficiency.} The latter is a~prerequisite for political and economic performance.\pozn{\fntit{Ibid}.} Furthermore, administrative procedure is nowadays the fundamental business process of public administration and aims to improve the rationality of its functioning through the alteration of statutes regulating administrative procedure, among other things in the sense of eliminating administrative barriers to the increased effectiveness of public policies.\pozn{Kovaè (2011).} Despite tendencies towards a~more partner-oriented relationship between the state and citizens (for example, mediation, administrative contracts, public-private partnership, etc.), the main, distinctive feature of state-citizen relations still remains the superiority of public interest over private interests. The functioning of the state and especially of its executive powers is regulated by administrative law (substantive and adjective). As part of its formal aspect, administrative procedure is a~tool for the creation and implementation of public policies in the development of states. The trend goes from strict authoritative functioning to the modernization of administrative procedures, i.e. partnership. The modern state is therefore more oriented towards the user of public services and being a~partner with the individual party\pozn{Schuppert (2000).} within the framework of the protection of public interest and the interest of third parties. At the end of the 1980s, several private-sector principles and methods were introduced in the public sector as a~result of the evolution of new public management (NPM) which emphasized the efficiency of management, democracy and user orientation. NPM principles were put into force to a~different extent in states and include privatization, decentralization, deregulation, new forms of responsibility and efficiency measurement. New trajectories have been identified in terms of: 1) NPM (with elements such as a~shift from internal towards external orientation towards meeting citizens' needs, performance indicators and market type mechanisms, decentralization and privatization, etc.); and 2) the Neo-Weberian state (NWS, with elements such as the reaffirmation of the state and representative democracy, preserving basic principles of administrative law, impartiality and hierarchy in the public service as a~distinctive function in society) -- even as a~global convergence.\pozn{Pollitt and Bouckaert (2011).} As a~theory and practice, NPM in particular and NWS were a~key source for the \fntit{doctrine of good governance} or New Public Governance (NPG) concept. According to this doctrine, the state should only ensure authority and the protection of general social benefit, but the state itself is not the only holder of this authority (development from the authoritative, centralized to the service, decentralized function of state, partnerships and networks). The good governance concept derived mainly from the private sector; it was introduced in 1990 by the World Bank and the emphasis at the time was on economic aspects (the connection between the quality of the system of state rule and the capacity to achieve sustainable economic and social development), while democratic aspects were not involved. Later, the concept was taken up by different international organizations and came to include democratic aspects.\pozn{Venice Commission (2011), cf. Pollitt and Bouckaert (2011).}

A trend of change from the top-down, unilateral behavior of public authorities in their relation to citizens to a~more horizontal relationship with characteristics of a~client relationship\pozn{The development of partner-oriented public administration was also a~result of a~critique of Weber's more or less rigid model of hierarchically organized and legally bound public administration.} (as in the private sector) and new values in administrative decision-making/the regulation of administrative services, such as transparency, simplicity, clarity, participation, responsiveness and towards citizen-oriented performance\pozn{Rusch (2011).} can be detected in recent decades. Another attempt at a~less top-down approach was the adoption of the European Commission's White Paper on European Governance\pozn{European Commission (2001). The definition given in the White Paper: \fntit{``Governance means rules, processes and behaviour that affect the way in which powers are exercised at European level, particularly as regards openness, participation, accountability, effectiveness and coherence.''}}, which concerns the way in which the EU uses the powers given to it by its citizens and establishes five principles of good governance: openness, participation, accountability, effectiveness and coherence. The main goal was the greater involvement of citizens/organizations in EU policy making. As stated in the white paper, these five principles are important for establishing more democratic governance and supporting democracy and the rule of law in member states; furthermore, they apply to all levels of government (global, European, national, regional and local).\pozn{\fntit{Ibid}.} According to SIGMA\pozn{SIGMA (1999, hereinafter referred to as SIGMA). Specifically on transparency on openness in Savino (2010).}, the key components of good governance are: rule of law; the principles of reliability, predictability, accountability and transparency; technical and managerial competence; organizational capacity and citizen participation. \fntit{Principles of public administration} were derived from these key components; they were further \mbox{defined} through the jurisprudence of national courts and the European Court of Justice and are now seen as \fntit{public administration standards}. According to SIGMA\pozn{\fntit{Ibid}.}, the main administrative principles common to Western European countries can be divided into four groups: legal certainty; openness and transparency; accountability; efficiency and effectiveness. Especially recently, increasing emphasis was also placed on the rationality and efficiency of administrative procedures in Eastern Europe, either in terms of implementing public interest or the mission of administrative bodies or in terms of, e.g. economic development, mitigation of economic crisis, improving investment capacities, etc.\pozn{Rusch (2009).} Therefore, a~good administration as a~well-functioning bureaucracy is one with the capacity to support the government and its partners in steering the society and the economy toward collective goals, being democratic, pursuing the rule of law and accountability in the broadest sense. 

The right to good administration as a~narrower, legal concept based on procedural rights is part of good governance doctrine.\pozn{Venice Commission (2011).} The concept of good administration can already be detected in Resolution (77)\pozn{Mendes (2009).} on the Protection of the Individual in Relation to the Acts of Administrative Authorities\pozn{Adopted by the Committee of Ministers on 28.9.1977.}, which regulated the right to be heard, access to information, assistance and representation, statement of reasons and indication of remedies. In 2007, on the basis of the European Code of Good Administrative Behaviour (2001) and Resolution (77)31, Recommendation CM/Rec(2007)7 of the Committee of Ministers to member states on good administration was adopted. The Recommendation contains a~Code of Good Administration (Code); it takes the view that good administration is one of the aspects of good governance and instructs member states (not just the EU, but the wider circle) to promote good administration within the principles of rule of law and democracy. The Code sets rules regarding the issuing of administrative decisions, appeals and compensation and promotes nine principles of good administration: lawfulness, equality, impartiality, proportionality, legal certainty, taking action within a~reasonable time limit, participation, respect of privacy and transparency.\pozn{In Continental Europe, the principle of transparency was initially implemented only by means of APAs (thus as ``procedural'' transparency), during the last two decades; however, procedural transparency has been complemented with the spread, all over Europe, of Freedom of Information Acts (Savino (2010)).}

Furthermore, the Charter of Fundamental Rights of the European Union was adopted in December 2000, initially as a~Recommendation and later, in 2010, as an obligatory primary law.\pozn{OJ C 83/389, 30.3.2010.} In Article 41, it regulates the right to good administration as one of citizens' fundamental rights. The right originally derived from the case-law of the Court of Justice of the European Union and the General Court and was initially binding as a~fundamental legal principle.31 According to Article 41, it demands the impartial, fair handling of affairs within a~reasonable timeframe and includes the right to be heard, the right to have access to files, the obligation to give reasons for a~decision, the right to use official EU languages and the right to compensation for damages caused by institutions/their servants while performing their duties. To explain the meaning of the right to good administration in practice, the European Code of Good Administrative Behaviour (European Code) was adopted in 2001 through a~resolution of the European Parliament. The right to good administration is today a~fundamental right consisting of a~set of individual rights and obligations aimed at preventing arbitrary administrative behavior in the European Union. The European Code represents the concretization of these individual rights and obligations.\pozn{Mendes (2009), Hirsch-Ziembiñska (2007).}

In Table\;1, some of the main international documents that provide guidelines on public administration (PA) for national systems' performance are analyzed, with a~focus on the concepts of good governance (GG) and good administration (GA). From the elements included in the concept of GG, we can confirm the assertion that this concept is of a~broader nature and also includes political dimensions linked to economic growth and social resources.\pozn{Venice Commission (2011).} The basic principles included in GG are more abstract and general (coherence, for example) and contain not only procedural requirements, but also substantive requirements explicitly demanding human rights protection. Furthermore, this concept involves more managerial principles; besides transparency, participation, effectiveness and efficiency, which are common to both concepts, technical and managerial competence, organizational capacity, accountability, simplification of procedures and responsiveness to people's needs are addressed. On the other hand, the concept of GA promotes the procedural aspects of PA performance to a~greater extent, i.e. use of language, the right to be heard, assistance and representation, reasoning and indication of legal remedies. It also includes compensation for damages, which partly overlaps with ethical behavior, combating corruption and accountability, all of which are typical of GG. 

Naturally, the concept of GA is more orientated towards human rights protection than the concept of GG, since the former notion derives from the \fntit{Rechtsstaat} doctrine while the latter promotes effective policy-making as well. However, the Venice Commission research\pozn{\fntit{Ibid}.} of 2011 explicitly included in the GG concept the presumption that GG also encompasses GA as a~narrower, legal concept, with the protection of human rights as one of its typical independent elements. As indicated in Table 1, certain key principles overlap and are basically the same, for example, rule of law includes the principles of lawfulness, impartiality, proportionality and certainty; equity is connected to equality, fairness and non-discrimination. Furthermore, both concepts promote participation and decision-making in a~reasonable timeframe, and these are inextricably connected to the principles of efficiency and effectiveness included in both GG and GA. There is, however, no one-way definition of either concept, meaning that their content can vary between different stakeholders. Nonetheless, the data presented in Table 1 include a~set of the most common contents of the concepts of GG and GA and should serve as a~model for setting up fundamental principles of administrative procedures. 

\clearpage\newpage

\popis{Table 1: Good Administration \& Governance Concepts as a~Basis for establishing Principles of Administrative Procedures}
{\footnotesize\tabcolsep 4pt\zactab{|p{20mm}|p{28mm}|p{15mm}|p{50mm}|}\hline
\bf Concept 	&\bf	Selected Documents 	&\bf	Issuer 		&\bf	Principles:\\
			&						&\bf	Year			&\bf	a) Classical constitutional safeguards\\
			&                                             &                     	&\bf b) \bf\fntit{    Managerial and modern NPM and }\\
			&						&				&\bf\fntit{    NPG/GG/GA principles}\\\hline
\bf\fntit{Good}			&	European Convention on 	&	Council of 	&	a) \\
\bf\fntit{Administration} 	&	Human Rights 			&	Europe 		&	1. Principle of lawfulness; \\
			&						&	1950		&	2. Equality, fairness, non-discrimination;\\\cline{1-3}
\bf\fntit{Good} 		&	Resolution (77)31 on the 	&	Council of 	&	3. Impartiality, proportionality, certainty, \\
\bf\fntit{Administration} 	&	Protection of the 		&	Europe 		&	4. The right to be heard, access to \\
			&	Individual in Relation to 	&	1977		&	information, respecting privacy;\\
			&	the Acts of Adm. 	 	&				&	5. Use of language; \\
			&	Authorities 			&				&	6. Assistance and representation;\\\cline{1-3}
\bf\fntit{Good} 		&	European Code of 	 	&	European 		&	7. Reasoning and indication of \\
\bf\fntit{Administration} 	&	Good Administrative  	&	ombudsman 	&	remedies; \\
			&	Behaviour				&	(EU) 	 	&	8. Compensation of damages.\\
			&						&	2001, 2005	&	b) \\\cline{1-3}
\bf\fntit{Good} 		&	Recommendation 		&	Council of 	&	1. Participation;\\
\bf\fntit{Administration} 	&	CM/Rec(2007)7 to 		&	Europe 		&	2. Transparency;\\
			&	member states on good 	&	2007 		&	3. Effectiveness, efficiency, taking action\\
			&	administration 			&				&	within a~reasonable time limit.\\\hline
\bf\fntit{Good} 		&	European Principles for 	&	SIGMA 		&	a) \\
\bf\fntit{Governance} 	&	Public Administration 	&	(OECD) 		&	1. Human rights protection;\\
			&	(Paper No. 27) 		&	1999 		&	2. Rule of law; \\
			&						&				&	3. Reliability and predictability;\\
			&						&				&	4. Coherence;\\
			&						&				&	5. Equity;\\
			&						&				&	6. Ethical behavior;\\
			&						&				&	7. Combating corruption;\\
			&						&				&	8. Decision-making in reasonable time.\\\cline{1-3}
\bf\fntit{Good} 		&	White Paper on 		 &	European 		&	b)\\
\bf\fntit{Governance} 	&	European Governance 	&	Commission 	&	\fntit{1. Transparency, openness, participation;} \\
			&	  					&	(EU) 		&	\fntit{2. Technical and managerial competence;} \\
			&						&	2001		&	\fntit{3. Organizational capacity;} \\
			&						&				&	\fntit{4. Accountability, effectiveness, efficiency;}\\
			&						&				&	\fntit{5. Simplification of procedures;} \\
			&						&				&	\fntit{6. Responsiveness to people's needs.} \\\hline
\bf\fntit{Good} 		&	Stocktaking on the 		&	Venice 		&	\fntit{Combining all GG \& GA principles.}\\
\bf\fntit{Governance} 	&	Notions of GG and GA 	&	Commission 	&	\\
\bf\fntit{and Good} 		&	(Study no. 470/2008, 	&	(Council of  	&	\\
\bf\fntit{Administration} 	&	CDL-AD(2001)009) 	&	 Europe) 		&	\\
					&						&	2011 		&	\\\hline
\kontab}
\zdroj{Source: selected documents as deriving from Table 1}

The rights to be exercised in administrative procedures are typified by their positive status, meaning certain activity is also expected of the parties (for example, to receive social assistance in cash, the party needs to submit an application to the respective public authority). Parties are bound to and dependent on the state/local authorities in enforcing their rights and also when authorities impose obligations upon them; consequently, respect for human rights is of great importance in these procedures as well. Most countries (Austria (1925), Yugoslavia (1956), Poland (1960), Hungary (1957), Czech Republic (1967, 2004), Germany (1976), Italy (1990), Slovenia (1999), Croatia (2009), Macedonia (2005)) adopted a~statute governing administrative procedure, and this serves as a~tool for the self-limitation of authority.\pozn{Statskontoret (2005).} Others developed rules on administrative procedure through case law and different parts of legislation without having a~statute explicitly regulating administrative procedure (e.g. Belgium, France, United Kingdom, Ireland).\pozn{Cf. Rusch (2009), Statskontoret (2005).} Regulation is aimed at protecting parties from the abuse of power in decision-making about administrative and other public matters. Regulation of this kind is a~sign of a~democratic regime and is extremely important for the protection of the individual's position, since it enables the predictability of public authorities' conduct. By means of administrative procedures, parties have several procedural safeguards which guarantee democratic procedures and the implementation of constitutionally ensured safeguards against the abuse of power. The main principles for enabling the protection of human rights and fundamental freedoms in administrative procedures are the separation of powers, legality, and equality before the law and proportionality.\pozn{Grafenauer and Breznik (2009).} These principles are usually already enshrined in the constitution. In general, legal principles are of great importance for all branches of law, administrative law being no exception. They present value-based criteria on how to behave in different legal relations, including administrative-legal ones. Principles give decision-makers the possibility of evaluating and estimating the relevance and use of legal rules in concrete case situations.\pozn{Pavènik (1997).} Rules define ways of behaving and acting, in contrast to principles, which do not have such a~distinctive one-way definition of behavior.\pozn{For more about theory on principles, see Ávila (2007).} In accordance with the principle of legality, the authorities are strictly bound to the legislation in force, which is abstract and general so as to achieve the widest possible circle of addressees. It would be irrational to expect all life situations to be covered by legislation; additionally, within the existing legal rules, decision-makers often come across situations where the applicability of a~certain rule is not clearly defined. Fundamental principles are the core interpretative instrument to be used in such cases.\pozn{More in Pavènik (1997).} Although legal principles are in a~sense ``blank concepts'' and, as such, somewhat elusive, theory and practice recognize them as being very important for lawmaking and enforcement.\pozn{SIGMA (1999).} Theoreticians like Dworkin and Esser see in principles not just an instrument to be applied when we need to define the use of a~certain legal rule, but also legal guidance which operates on the same level as legal rules and has a~direct impact on everyday judicial decision-making.\pozn{Novak (2010).} Principles are no longer viewed as being of subsidiary relevance, but must be used by judges simultaneously with rules.\pozn{Cf. ibid.}

\clearpage\newpage

\podnadpis{Regulation of AP in selected Eastern European States}
\vspace*{2mm}
\oddil{The Enactment and Scope of AP in Selected States}
\vspace*{3mm}
Taking into account their common historical background and geographical position, we decided to compare the regulations on administrative procedures in the Republic of Slovenia, the Republic of Croatia, the Former Yugoslav Republic of Macedonia and the Czech Republic. These states represent a~very interesting combination of legal systems influenced by both western and eastern approaches. They were all influenced by the Germanic legal system, particularly the Austrian legal model,\pozn{According to Schwarze (1992), Ziller (2009, cop. 2007) and Statskontoret (2005), in terms of good administration four traditions of administrative law may be identified in Europe (and more broadly): 1) the administration-centered tradition, as in France, 2) the individual-centered tradition, as in the UK, Ireland, and the US, 3) the German-Austrian legislator-centered Rechtsstaat, and 4) the ombudsman-centered tradition, as in Scandinavia.} and also by the socialist regime following World War II. We could say that history partly repeats itself, in the sense that these states were all part of some kind of federation in the past, strived for independence in the 1990s, and are now coming ``back together'' under the wings of the European Union: Slovenia and the Czech Republic have been members since 2004, Croatia joined in July 2013 and Macedonia is a~candidate country. All these states are (being) Europeanized in a~sense as they adopt certain European features.\pozn{Petrov and Kalinichenko (2011). See also Olson (2002), and Radaelli and Featherstone (eds.) (2003).} National reactions to EU ``pressures''\pozn{Iancu (2012).} resulted \fntit{inter alia} in changes of the administrative systems of these states (so-called administrative Europeanization)\pozn{Hadjiisky (2009).}. Organization of national administration is in principle not part of \fntit{acquis} but the White Paper of 1995\pozn{European Commission (1995).} stressed the importance of administrative capacity for implementing the \fntit{acquis}. As a~result, several reforms of the administrative system of these states took place. Individual reforms in Slovene public administration and the whole series of activities carried out in this regard may be considered successful if the criteria for assessment are the objectives set in the reform documents and comparative indicators in the EU or other countries based on good administration and governance concepts. However, reforms were designed and carried out rather legalistically, with no functional analysis and a~lack of political coordination.\pozn{Kovaè (2013).} In Croatia, public administration reform has gone through numerous reforms; nevertheless, the main priority at all levels has always been the decentralization of tasks and structures. Major progress has been made in the area of legislation in order to improve public administration efficiency.\pozn{Nikolov (2013).} Macedonia has gone through a~period of complex economic, political and administrative development since gaining independence in 1991. As a~result, a~number of strategic and action plans have been adopted and implemented. One of the greatest concerns has been the high ``politicization of the public service''. In December 2010, the new Strategy on Public Administration Reform (2010--2015) was adopted and the main results are expected to be in the areas of public finances, HRM, e-government and management and corruption.\pozn{\fntit{Ibid}.} The reform of public administration in the Czech Republic started immediately after the Velvet Revolution of November 1989. The central state administration reform started many years after the public administration reform on local and regional tiers was completed and the reform processes and goals were not only because of the admission of the Czech Republic to the EU.\pozn{Vidláková (2006).} The reform process is far from being completed and many of the important reforms introducing modern approaches to public administration have been started only recently.\pozn{Hladík, Kopecký (2013).}

As far as administrative procedure is concerned, all four states have an Austrian administrative procedure regulation model in common. Namely, Austria adopted its law on administrative procedure at an early date, in 1925, and was followed by Czechoslovakia in 1928 and the Kingdom of Yugoslavia in 1930. The beginnings of the administrative procedure regime in the territory of Slovenia date back even further, to 1923. In 1946, the first law on general administrative procedure in Yugoslavia was abolished, and a~new federal law was enacted in 1956.\pozn{Kopriæ (2005).} In the interim period, procedural provisions of different laws, and partly also of sectoral laws, were applied. During the over 40 years it was in force, the federal law was amended only four times (in 1965, 1977, 1978 and 1986). 

The Republic of Slovenia became independent in 1991 and succeeded the legal order of the former Yugoslavia to an extent which was in accordance with the new state regime. The General Administrative Procedure Act (GAPA) was therefore taken over as a~``Slovenian law''. Only in 1999 did the Republic of Slovenia adopt a~new act (which entered into force in 2000); it was still based on the Yugoslavian one but introduced certain changes (the use of APA also for public service providers when deciding on rights/obligations; the subsidiarity principle; redefinition of legal principles; reduction of legal remedies, etc.)\pozn{More in Kovaè (2011).} with the aim of adapting the regulation to the new organization of the state and the new market-driven societal environment. In just 14 years, the Slovenian GAPA has already been amended nine times (see Table 2) with a~focus on the need to reduce administrative burdens and the introduction of new technologies (the exchange of data from official records by civil servants, e-communication, renunciation of the right to appeal, the right to access to files and access to public information, etc.). Its volume is still extensive, with 325 articles. With a~view to reducing administrative burdens, a~new amendment entered into force in March 2014. The main anticipated change is the demand that officials fill out a~form instead of parties. The authors are slightly skeptical about the actual need for and benefits of such changes and also feel that they bring up questions of the liability of officials, in the sense that new disputes could arise between the authorities and parties.

\clearpage\newpage

The Republic of Croatia also became independent in 1991 and took over the Yugoslav GAPA. In April 2009, a~new GAPA was adopted, and it entered into force on 1 January, 2010.\pozn{The first draft of the new GAPA was prepared as a~result of the CARDS 2003 project: Support to the Public Administration and Civil Service Reform.} To date (in the last three years) there have been no amendments. Croatia certainly took longer than Slovenia to enact a~new GAPA, but this "delay" resulted in far more new, modern approaches and half the articles. Some of the novelties are the introduction of new or certain amendments of existing principles; the regulation of administrative contracts; the ``silence means consent'' institute (if explicitly foreseen by the law); the amendment of the legal remedy system (the introduction of objection for administrative contract disputes and actions of public authorities and public service providers); e-communication, etc. In view of the APA's novelties, the amendment of sectoral regulation is now needed as an alignment process, especially since the new APA does not explicitly define its subsidiary application.\pozn{Hus and Katiæ Buba¹ (2011).} The Croatian GAPA definitely took an interesting approach, combining traditional regulation (such as the Yugoslavian GAPA) and modern approaches.

The Former Yugoslav Republic of Macedonia became independent in 1991 and also took over the Yugoslavian GAPA of 1956. In 2005, the Macedonian Law on General Administrative Procedure (LGAP) was enacted; it has since been amended twice, in 2008 and 2011. The LGAP of 2005 was more or less the continuation of the Yugoslavian GAPA with a~few minor changes, such as new principles and delivery methods.\pozn{Davitkovski and Pavlovska-Daneva (2009). Public Administration Reform Strategy 2010--2015, Republic of Macedonia, 2010, hereinafter referred to as Public Administration Reform Strategy 2010--2015.} The main novelties introduced through the 2008 amendments were the shortening of deadlines for decisions in administrative matters (15 days for simple matter (previously one month) and 30 days for complex cases (previously 60 days)) and the removal of the two-level decision rule; the right to an appeal became a~possibility that can be prescribed by the GAPA or substantive laws\pozn{Public Administration Reform Strategy 2010--2015. This amendment was made to implement constitutional change, which allowed an appeal or \fntit{other legal protection} against first-instance administrative acts, meaning that a~direct administrative dispute against first-instance decision was possible (Davitkovski and Pavlovska-Daneva (2009)).} and the ``silence means consent'' mechanism was introduced.\pozn{Todevski (2011).} The 2011 amendment further developed this mechanism and defined civil servants' obligation to gather all documents from official records. To introduce the ``silence means consent'' principle, the Macedonian Parliament amended over 130 laws. However, the complex appeals system still undermines the application of this principle.\pozn{European Commission (2012). For the functioning of the ``silence means consent'' mechanism, see Todevski (2011), Davitkovski and Pavlovska-Daneva (2009).} One of the priorities of the Macedonian government is the modernization of the public administration system, within which administrative procedure reform is one of the most important parts. Administrative procedures are still seen as too lengthy and unnecessary.\pozn{Todevski (2011).} In 2010, the ``Public Administration Reform Strategy 2010--2015'' was adopted, and in March 2012 the European Commission and Macedonian government launched the High Level Accession Dialogue (HLAD), which is aimed at promoting the EU accession reform process, with public administration reform being one of five priority policy areas. One of the results of the first HLAD session was the government's decision to elaborate a~new Law on General Administrative Procedure (LGAP), since a~good LGAP is a~precondition for quality and legally correct administrative decisions; administrative procedures are in turn important for good administrative behavior, serving the community, promoting social trust in the executive branch, political stability, social wealth and economic development.\pozn{SIGMA (2012).}

The disintegration of Yugoslavia was followed in 1993 by the splitting-up of the Czechoslovak Federal Republic, whereby the Czech Republic became independent and adopted a~new constitution (already in December 1992). Part of Czech constitutional law\pozn{Having the same force as the constitution (see Pouperová, 2014).} is the Charter of Fundamental Rights and Basic Freedoms, which was taken over from the former legal order. Since 2004, the Czech Republic has been a~member of the European Union. Similarly to the ex-Yugoslav states, the development of public law in the Czech Republic was strongly influenced by Austrian law and after 1950 also by the Soviet system. After 1989, important changes were initiated, with an emphasis on fundamental human rights and freedoms.\pozn{Sta¹a and Tomá¹ek (2012).} Administrative procedure was codified in Czechoslovakia in 1928, also following the Austrian codification of 1925. Administrative procedure was enacted in 1967, and was amended several times, especially during the 1990s.\pozn{Statskontoret (2005).} In 2004, a~new Administrative Procedure Code was enacted (it entered into force in January 2006) with the aim of simplification.\pozn{Besides this Code, there is also other legal regulation on administrative processes that deals with different activities of administrative bodies. For more on these activities and about particular regulation of administrative procedure, see Sta¹a and Tomá¹ek (2012). Furthermore, the tax and charges regulation is completely autonomous, meaning the administrative procedure code does not apply. Slovenia, on the other hand, enacted a~special Act on Tax Procedure in 2006, which has been amended several times and has a~total of 424 articles; nevertheless, the General APA is still applied in accordance with the principle of subsidiarity.} The nature of the code is subsidiary, and it was followed by more than one hundred special regulations. As can be seen below in Tables 2 and 3, the Code takes certain modern approaches, such us public contract regulation, issuance of measures of a~general nature, complaints against inappropriate conduct of persons in authority, etc. 

From the comparison shown in Table 2, we can conclude that there are certainly commonalities among the APAs of all the selected states. Typical of all four is the scope of APAs for issuing individual concrete administrative acts when public authorities (state, local/regional authorities and other subjects when partaking in authoritative decision-making) decide in the field of administrative law. Moreover, it is directly evident that more recent laws (even when it is a~matter of just a~couple of years) incorporate more modern approaches with a~broader scope, including agreements and administrative contracts and the general design and implementation of administrative acts; Croatia and the Czech Republic are role models for this approach. The APAs compared here also simultaneously regulate these procedures on a~more abstract level and seek to provide principles for (overly) detailed operational rules. Such a~flexible approach allows for the greater incorporation of a~broader spectrum of life events under the regulation and does not require constant amendment, which could decrease legal certainty and the transparency of the regulation. 

\popis{Table 2: The Enforcement and Scope of APAs in Selected States}
{\footnotesize\zactab{|l|l|l|l|}\hline
\bf State 		&\bf	Enforcement 	&\bf	No. of Art. 	&\bf 	Scope of Application/Regulation\\
			&\bf	/Codification 	&				&	\\\hline
\parbox[t]{19mm}{\bf Slovenia} 		&	\parbox[t]{22mm}{\raggedright 2000 / 1999\\Am.: 2000, 2002, 2004, 2005, 2006, 2007, 2008, 2010, 2013} 	&	\parbox[t]{12mm}{325}			&	\parbox[t]{54mm}{Issuing individual administrative acts (deciding about concrete administrative-legal relations); use of GAPA when deciding on other public matters for which special procedures have not been enacted\footnotemark{} (Art. 1--4).}\\\hline
\parbox[t]{19mm}{\bf Macedonia}  	&	\parbox[t]{22mm}{2005 / 2005\\Am.: 2008, 2011} 	&	\parbox[t]{12mm}{302} 	&	\parbox[t]{54mm}{Issuing individual administrative acts in administrative matters (i.e. all acts and activities through which public administration authorities are represented and executed) (Art.\;1 and~2).}\\\hline
\parbox[t]{19mm}{\bf Czech Republic} 	&	\parbox[t]{22mm}{2006 / 2004}	&	\parbox[t]{12mm}{184} 	&	\parbox[t]{54mm}{The process of administrative authorities\footnotemark{} when issuing decisions to constitute, alter or cancel rights and/or duties of particular persons in particular cases/issuing decisions declaring that a~person does not have rights/duties in a~particular case (Sec. 9). Contentious proceedings, disputes arising from public contracts, disputes arising from civil, labor, family or business legal relations (only when provided by special statute) (Sec. 141). Proceedings to determine a~legal relationship (Sec. 142). Issuing statements, certificates, notifications (Sec. 154--158). Public contracts (Sec. 159--170). Issuing measure of general character (Sec. 171--174).\footnotemark}\\\hline
\parbox[t]{19mm}{\bf Croatia} 	&	\parbox[t]{22mm}{2010 / 2009} 	&	\parbox[t]{12mm}{171}	&	\parbox[t]{54mm}{Issuing individual administrative acts; administrative contracts; for the protection of the rights/legal interests of parties when public service providers decide on their rights/obligations/legal interests\footnotemark; any other activity of public administration bodies in the field of administrative law that directly affects the rights, obligations or legal interests of parties (Art. 1--3).}\\\hline
\kontab}
\zdroj{Source: APAs of compared states}

\footnotetext[68]{Minimal procedural safeguards should be ensured also in these matters (e.g. misdemeanors, disciplinary procedures, etc.).}
\footnotetext[69]{Authorities of the executive branch of the government, of regional self-governing units and other authorities when performing competences in the field of public administration (see Section 1).}
\footnotetext[70]{In general, the APA does not apply to civil, business, or labor legal actions performed by administrative authorities, etc. (Section 1/3). For the scope of application of the Czech APA, see also Sta¹a and Tomá¹ek (2012): the current concept of administrative procedure regulation in the Czech Republic is focused on \fntit{activities of the administrative body} and not on the persons to whom the conduct is directed. The issuing of measures of a~general character is only one instance of procedure; the measure can be subject to judicial review by administrative courts (Sta¹a and Tomá¹ek (2012)).}
\footnotetext[71]{If no judicial or other legal protection is prescribed by law (see Hus and Katiæ Buba¹ (2011): a~school's decision on children's diet, a~doctor prescribing incorrect treatment to a~patient).}

\clearpage\newpage

\oddil{Notions of European convergence in Eastern Europe}
Over the course of history, each country has developed its own administrative system within its legal tradition, which is why the EU allowed differences in the administrative legal systems of its member states. Convergence developed through different acts (conventions, recommendations, declarations, and directives) that prescribe general principles, standards of public administration (reliability, predictability, openness, transparency, responsibility, efficiency, economy, effectiveness, participation of citizens, etc.)\pozn{Davitkovski and Pavlovska-Daneva (2010).}, based, among other things, on the common legal tradition and democracy of member states. Therefore it is not surprising that to date there has not been a~coherent codification of administrative law for the EU, but only some fragmented regulation. A~special role in the field of shaping (European) administrative law has been given to the case law of the Court of Justice. From it, legal principles governing administrative activity in European Union law have been derived. In its case law, the court relied on the unwritten general principles common to the constitutional/administrative traditions of member states.\pozn{Schwarze (2011).} However, ``European administrative law'' is a~relatively new branch of law; in a~narrow sense it represents administrative law for the EU, and in a~broader sense the approximation of the administrative laws of member states on the one hand and of the EU on the other.\pozn{\fntit{Ibid}.} The core of European administrative law is a~combination of principles of legality, proportionality, and legal certainty, protection of legitimate expectations, non-discrimination and fair administrative process and effective judicial review.\pozn{Schwarze (2011).} 

The comparative overview of the selected regulations revealed that the administrative procedures share a~large number of common institutes. This is to be expected, since in all four states the codification of administrative procedures has Austrian roots (the law of 1925 being a~model) and since they all fall under civil system law and the Germanic continental legal space. Furthermore, all the states decided to join the European Union, with Slovenia and the Czech Republic joining in 2004, Croatia joining in 2013 and Macedonia being a~candidate country. The harmonization of a~national legal system with the EU \fntit{acquis communautaire} is one of the preconditions for successful accession to the EU. 

In comparison to other EU countries, all four states examined here have a~very long tradition and history of administrative procedure codification. Namely, most EU member states enacted (or modified) administrative procedure only after the 1990s.\pozn{Statskontoret (2005).} This is definitely a~result of societal changes and not only reflects the fundamental need for the protection of the public interest and the protection of individual rights, but also pronounced partnership tendencies which require greater flexibility in public decision-making. Administrative procedure is especially important for different segments of business processes (such as establishing a~company, taxes, public-private partnership, etc.); this is particularly true of the EU, where members' free market rights also influence countries' economic growth. Although this long tradition of AP codification is a~sign of a~democratic approach to regulation and the rule of law, it can also lead to some rigidity and inflexibility. That is not entirely bad, since only copying ideas and solutions from other (especially western) countries' legal systems is not necessarily the most effective approach to change. A~country's legal tradition and history is a~very important part of its ``legal identity'' and intrusive, alien interventions can produce even less effective conduct or become mere nice-looking legal institutes on paper that are never used. Both are ultimately unproductive. Each change in regulation should therefore be carefully thought through and multidisciplinary research (from at least a~legal and economic perspective) should be conducted beforehand. It does seem -- from a~subjective standpoint, of course -- that most Eastern European countries are rushing headlong into the European Union, taking over its legal order without much resistance for the sake of successful accession. Most of these countries are still young democracies in terms of the length of the period in which they have been independent; thus certain deviations in the effectiveness of public administration activities are normal. Namely, the process of evolving a~legal order is also affected by society itself, especially by its values and ``ripeness''. Therefore the enhancement of the public administration system is an ever-evolving process in each country, and as such requires its own time and space. The Croatian and Czech APAs are definitely already taking modern approaches with new institutes, such as public contracts, the regulation of abstract acts, etc. With the new APA it is preparing, Macedonia has a~great opportunity to enhance its regulation. Until now, Slovenia was satisfied with the APA of 1999, although there has been a~tendency to patch it up every few years with certain novelties. Yet there are murmurings of the adoption of a~completely new law which could possibly take the simplification approach (a shorter law with less formalistic regulation and greater flexibility), although the questions to be asked here are how much discretion will be given to public administration and is society, in light of its low degree of trust in the rule of law, ready to put a~greater degree of trust in public authorities. In the authors' view, one of the main guidelines for possible changes in any kind of regulation in any state in the future should be to enact clear regulation that is understandable to the average population (keeping in mind the simplicity of procedures) and marked by continuation (i.e. not subject to change with each change of government -- this applies especially to sectoral regulation).
   
\podnadpis{Principles in Administrative Procedure Acts as a~Way to Ensure Human Rights}
Principles are the fundamental value base of any legal regulation to be followed in its design and interpretation in any legal procedure.\;Administrative procedure is the main process of public administration and an instrument of democracy aimed at protecting (interested) parties in their interactions with authoritative decision-making. Administrative procedural law principles are a~traditional tool of democracy whose purpose is to make operational the basic constitutional safeguards for the protection of individuals\;from\;the\;abuse\;of\;power. Procedural guarantees are therefore important for the protection of fundamental rights and freedoms. Furthermore, the protection of fundamental, constitutional rights is inextricably linked to correct administrative procedure.\pozn{Schwarze (2004).}\;A\;basic\;guideline\;for the regulation of \mbox{administrative} procedures was given in 2005 by SIGMA, which stated \fntit{general principles of administrative activity} (such as legality, impartiality, procedural fairness with timeliness, openness and transparency, accountability and liability, scope of the law and interested parties) and \fntit{principles governing administrative procedure}: definition of administrative activity, form of AP (general and special regulations of AP), the commencement of AP, language, evidence, hearing of the interested parties, time limits and reinstatement, termination, administrative silence and retrial. However, certain fundamental principles are usually already defined by the constitution and not always repeated in GAPA. The latter of course does not diminish their legal validity.\pozn{Cardona (2005), Statskontoret (2005).} The overview of the APAs compared here showed that the majority of the abovementioned (general) administrative principles as set by SIGMA in 2005 are present, some as fundamental principles, others as rules. 

Regulation of fundamental administrative principles is typically part of the introductory provisions of APAs, which apply to administrative procedure as a~whole. APAs or sectoral regulation can also define certain special principles which are used in certain parts/types of procedure. With the exception of the Croatian APA, the APAs examined here explicitly recognize the ``principle'' of subsidiarity (Czech APA: Section 1/2; Macedonian APA: Article 3; Slovene APA: Article 3), meaning provisions of the General APA, including principles, are also applied in special administrative procedures.\pozn{Cf. Statskontoret (2005).} Article 3/1 of the Croatian APA, on the other hand, explicitly requires the use of the GAPA in all administrative matters except in certain individual questions of administrative procedure which can be regulated differently in other statutes if this is necessary for certain administrative fields and not in contradiction with the GAPA (it provides no direct ordinance for the subsidiary use of GAPA).\pozn{Hus and Katiæ Buba¹ (2011).} Despite the lack of a~direct ordinance, such a~rule can still be derived from a~substantive definition of administrative matters that requires the use of GAPA whenever the authority decides on a~right/obligation or legal benefit of a~party in a~field of administrative law where the actual or at least potential existence of a~conflict between the public and private interest may occur.

In Table\;3, we identified two groups of principles: 1) classical, such as the notion of human rights protection and 2) managerial and modern, such as indicators of concepts of NPM, GG and GA. Such a~classification is based on overall public management and governance features, types of reforms and politico-administrative regimes, setting a~framework for administrative procedures as well. Namely, the trends of reforms are focusing, in different countries and time periods, on: 1) classical, \fntit{Rechtsstaat} and public interest protection based reforms, and 2) reforms, prioritizing NPM/NPG principles, such as effectiveness, performance orientation, transparency, accountability, etc.\pozn{Pollitt and Bouckaert (2011).} However, some of these principles are rather difficult to categorize as such, since they emerge in different perspectives and legal or policy frameworks and papers as a~classical safeguard or/and modern standard -- typically the principle of transparency or openness.\pozn{Savino (2010).} 

\clearpage\newpage

Nevertheless, for the purpose of this study we identified a~total of 13 principles in the classical group. Number one was the principle of legality, which in its broadest sense also includes principles of predictability, finality, impartiality and proportionality (see numbers 1--4 in Table\;3). These principles are an essential part of the rule of law and a~precondition of any democratic state, as they constitute a~procedural presumption for human rights protection. Continuing, classical principles ensuring participation of the interested parties are included in the first group, i.e. protection of affected persons' rights, assistance, hearing and right to defense, use of language, data access and protection and the right to legal remedy. Within these principles, only one duty is imposed upon the interested parties: the duty to speak the truth and the fair use of rights. Otherwise they impose obligations upon the authority so as to ensure the (human) rights protection of parties, which are in an a~priori weaker, subordinate position and as such need to be protected (for example, substantive truth and the timeliness of decision-making, see numbers 8, 10 and 12 in Table\;3).

In restructuring the vision of the authorial state and its functions, the need to introduce managerial practices into public administration has been an important objective in Eastern European countries since the 1990s -- most often because of external supranational incentives (EU, OECD, and World Bank). Given the specific socio-political and economic circumstances of that time, the primary goal of introducing NPM in those countries was in fact quite different from those pursued in older democracies, since the latter primarily see NPM as a~tool for upgrading democratic system making, while the former mostly applied it as a~tool for stimulating effective and efficient ways of a~democratic system in general, and only later on for modernization and adaptation purposes. Hence, we identified a~total of only four principles of a~managerial and modern nature in the Macedonian and Czech regulations. Both of these regulations promote public administration as a~service to the public, demanding polite behavior and the satisfaction of the needs of the affected persons when possible. The Macedonian APA also explicitly mentions the accountability of the authorities. The Czech APA goes even further, including the right to complaint about the inappropriate conduct of persons in authority or against the proceeding of an administrative authority (Section 175) and introducing the principle of the mutual cooperation of administrative authorities in the interest of good administration (Section 8/2). In theory, some of the basic principles of the Czech APA are seen as the projection of general legal principles onto positive law.\pozn{Sta¹a and Tomá¹ek (2012).} Some principles are derived from the Constitution and the Charter of Fundamental Rights and Freedoms, such as the principle of Section 2/4 (``the adopted solution should be in accordance with public interest, administrative authority complies with circumstances of the particular case and there should be no groundless differences when deciding factually identical/comparable cases''), which introduces the principles of predictability and stability in decision-making. Czech jurisprudence understands them as the principle of (protection of) legitimate expectations (predictability of a~decision) or as the principle of justified confidence in the procedures of public administrative authorities; they are also understood as the equality of addressees, the prohibition of the abuse of discretion and the requirement to reason a~decision.\pozn{Skulová, Husseini, Vrbová, and Prokopová (2011).} The principle of equality also derives from Section 7, which states the equal status of affected persons' procedural rights and demands the performance of procedural duties in the same manner from all affected persons. The principles of activities of administrative authorities as enshrined in Sections 2--8 correspond to a~considerable degree to the principles of good governance.\pozn{\fntit{Ibid}.} 

Based on the normative analysis in Table 3, a~number of similarities in all four APAs can be detected; on the other hand, there are also some differences. The principles identified as common to all four states are the principles of legality, equality, impartiality, objectivity/independence, proportionality, protection of public interest, protection of parties' rights, protection of the economy/effectiveness of procedures, ascertaining the case without reasonable doubt/substantive truth and the right to legal remedy (appeal).

\popis{Table 3: Principles in APAs of Selected States\pozn{The content of the table focused on the introductory provisions of APAs that regulate fundamental principles of administrative procedures. Although some of the principles are not strictly mentioned in these provisions and may therefore not be marked in Table 3, they might be defined in later APAs provisions or other regulation (like the constitution, the Public Administration Act, the Law on Free Access to Information of Public Sector, etc.). Table 3 does not include certain Slovene and Macedonian APA institutes regulated in the basic principles chapter, such as the definition of administrative matter, authorities' obligation to use the APA, definitions of authority/civil servant and subsidiary application of law. (Cf. also Hus and Katiæ Buba¹ (2011), and Skulová, Husseini, Vrbová, and Prokopová (2011)).}}
\vspace*{-4mm}
{\footnotesize\tabcolsep3pt\zacltab{|l|p{54mm}|c|c|c|c|}\hline
\bf No.	&\bf	Principles:									&\bf	Slovene 		&\bf	Macedonian 	&\bf	Czech 		&\bf	Croatian \\
		&\bf	1--13 Classical (protection of HR)				&	(Art. 6--14)	&	(Art. 4--19)	&	(Sec. 2--8,	&	(Art. 5--14)\\
		&\bf\fntit{14--17 Managerial and modern (GG and GA)}	&				&				&	81--, 175)	&	\\\hline
\endhead
\mc{6}{r}{\fntit{Continued on next page}}\endfoot
\endlastfoot
\01		&	Legality 	&\bf x&\bf x&\bf x&\bf x\\\hline
\02 		&\raggedright Predictability and stability of decision-making/ Protecting rights attained by the parties/ Finality/ Legal effectiveness of the decision
		&	&	\bf x&\bf x&\bf x\\\hline
\03 		&\raggedright	Equality, impartiality and objectivity/ Independence 	&	\bf x&\bf x&\bf x&\bf x\\\hline
\04 		&\raggedright	Proportionality in protection of parties' rights and public interest 	&	\bf x&\bf x&\bf x&\bf x\\\hline
\vrule width 0pt height 2.8mm \05 		&\raggedright Protection of affected persons' rights/ Assistance/ Equality &\bf x\footnotemark 	&\bf x&\bf x&\bf x\\\hline
\06 		&\raggedright Notification of the affected persons and right of defense/ Hearing &\bf x&\bf x&\bf x& \\\hline
\07 		&\raggedright Official use of languages and scripts / Alphabet 	&	&\bf x 	&	&\bf x\\\hline
\08 		&\raggedright Ascertaining the case without reasonable doubt/ Substantive truth &\bf x&\bf x&\bf x&\bf x\\\hline
\09 		&\raggedright Data access and protection 	&	&	&	&	\bf x\\\hline
10 		&\raggedright Independence/discretionary evaluation in assessing evidence &\bf x&\bf x 	&	&\bf x\\\hline
11 		&\raggedright Duty to speak the truth and fair use of rights/ Notify administrative authorities about on-going public procedures &\bf x &	&\bf x&\\\hline
12 		&\raggedright Bringing decisions without unnecessary delays/ 		&\bf x&\bf x&\bf x&\bf x\\
		&\raggedright Economy and urgency of the procedure/ Efficiency	&	&	&	&	\\\hline
\vrule width 0pt height 2.8mm 13		&\raggedright Usually two-instance administrative procedure when issuing individual decisions/ Right to appeal/Legal remedy\footnotemark 	&\bf x &\bf x\footnotemark	&\bf x&\bf x \\\hline
14 		&\fntit{Accountability} 	&	&\bf x 	&	&	\\\hline	
15 		&\raggedright \fntit{Mutual cooperation of administrative authorities in the interest of good administration} 	&	&	&\bf x 	&	\\\hline
\vrule width 0pt height 2.8mm 16 		&\raggedright \fntit{Public administration is a~service to public/ Service orientation} &		&\bf x &\bf x\footnotemark 	&	\\\hline
17 		&\raggedright \fntit{Right to complaint against inappropriate conduct of persons in authority or against the proceeding of administrative authority} &	&	&\bf x&\\\hline
\konltab}
\vspace*{-5mm}
\zdroj{Source: APAs of compared states}

\footnotetext[87]{Part of the principle of protection of parties' rights and public interest (Article 7).}
\footnotetext[88]{Final decision is subject to judicial review in all four states (cf. Sta¹a and Tomá¹ek (2012)).}
\footnotetext[89]{With the 2008 amendment, the right to appeal became only an option, not a~necessity -- being regulated by law, meaning a~two-instance rule is not basic principle of APA, although still enshrined in the chapter on Basic Principles (Davitkovski and Pavlovska-Daneva (2009)).}
\footnotetext[90]{The requirement of polite behaviour (see Section 4/1).}

Most of the common principles in the Slovenian, Croatian and Macedonian APAs can be viewed as a~consequence of the previous Yugoslavian APA (see Table 3). Macedonia also kept some principles which the Slovene and Croatian APAs excluded, such as accountability, final decision and the legal effectiveness of the decision. These institutes are not defined as general principles in the other countries' APAs, but they are derived from other provisions or other systemic regulation (for example, the Slovenian Civil Servants Act, Public Administration Act, etc.) or are already defined by the constitution (i.e. liability for damages, legal effectiveness (Articles 26 and 158 of Slovenian Constitution)). Croatia introduced some new principles in 2009, such as the principle of the right of parties to legal remedy (which includes multiple remedies: complaint, objection against public contracts or other activities of a~public body/public service provider and administrative dispute) and the principle of data access\pozn{Right of access to documents in an administrative procedure is an essential part of transparency regime. In all European legal orders, a~person participating in an administrative procedure has the right of access to documents on which the decision of the administration is (or will be) based (Savino (2010)) -- but not in all legal orders is this technically speaking formulated as a~general principle of APA.} and protection, which gives parties the right to access data, Internet pages, forms and professional advice/help. The principle of the hearing of the party (or any affected person with a~legal interest) is no longer explicitly expressed in general provisions, although it can still be derived from later APA provisions (for example Articles 30, 52). The Czech APA sets it as the right of defense (Section 4/3 and 4/4). The Slovene (Article 9) and Macedonian (Article 10) APAs took over the Yugoslavian principle of parties' right to express themselves before a~decision is issued. Another important dimension of parties' participation rights is the use of languages/alphabets/scripts, which is defined in all the regulations: in Macedonia and Croatia as part of the introductory provisions, in the Czech (Section 16) and Slovenian (Article 62) APAs as rules of procedure. 

\clearpage\newpage

Taking into account the subordinate position of the party, which puts the administrative procedure in the role of major protector of parties' rights when interacting with the authorities, the following set of fundamental principles as a~major interpretative instrument of rules were identified as an expression of parties' (human) rights protection: the principles of legality (in the broadest sense), equality, participation; protection of the parties' rights; the right to legal remedy (appeal); reaching decisions without unnecessary delays; and the principle of protecting rights attained by the parties (\fntit{res iudicata}). Together with the obligations imposed on the authorities (such as the principle of independence, establishing substantive truth, mutual cooperation of authorities in the interest of good administration, accountability, impartiality and objectivity), these provisions are also a~procedural safeguard and guarantee for the protection of human rights in administrative procedures, and in this respect they constitute administrative law as \fntit{materia constitutionis}.

Finally, the data collected in the survey (Table 3) confirm the initial hypothesis on the importance of administrative procedural law principles for the protection of human rights in the sense that these principles implement the operational safeguards already defined in the constitution. Namely, we have to bear in mind the importance of procedure for executing substantive rights of a~constitutional nature. For the authoritative acts governing these rights to be lawful, both formal and substantive legality has to be ensured. This means that the procedure itself has to be performed in accordance with legality so as to properly ensure substantive (human) rights. With their core interpretative function and constitutional background, fundamental principles are therefore of great importance for the protection of human rights and ensure legality and legal certainty also when deciding on administrative matters.

\podnadpis{Conclusion}
The research we conducted led us to conclude that AP regulation in the selected Eastern European states is still of a~classical (administrative law oriented) nature; it arises from a~subordinate citizen-authority relationship and protects the weaker party by prohibiting the arbitrariness of authoritative decision-making on the basis of the principle of separation of powers (two-instance procedures, administrative decisions being subject to judicial review). Therefore the purpose of APs and content deriving from analyzed principles confirms the hypothesis of APs' relevance and importance for the protection of human rights in administrative decision-making. Furthermore, comparative normative analysis of the APAs shows that the prevailing legal tradition has the Austrian APA as its basis, with certain new modern approaches such as public contracts, issuance of general acts, the institute of ``silence means approval'', public service providers' obligation to apply the APA when deciding about citizens' rights/obligations, e-communication, etc. Despite the shared historical heritage, differences appeared in the last decade, as Croatia and the Czech Republic introduced more managerial and modern institutes alongside classical approaches to AP regulation. However, on the basis of the three generations definition of administrative procedures\pozn{Barnes (2010).}, we can conclude that regulation in all four \mbox{countries} still belongs to the first generation of AP regulation, although certain citizen-state partnership elements can be noted. New or NPM/NPG-driven AP principles such as the mutual cooperation of authorities, accountability, the service orientation of authorities, etc., have been defined, and this reflects good administration approaches. Furthermore, all four countries' APAs include other main elements reflecting good governance (and also good administration), such as transparency, responsiveness, rule of law, effectiveness and efficiency.\pozn{Cf. Apelblat (2011).} Performance and respect for these principles of good administration as part of good governance are also a~prerequisite for the protection of human rights in practice, when implementing the regulation set. Finally, the greater a~country's awareness of the importance of AP for the protection of constitutional (human) rights, the greater its emphasis on building a~democratic state also through administrative procedure that introduces good governance and good administration principles. As a~result of (future) EU accession/membership (influencing the regulation also by funding different projects to encourage reforms, such as CARDS 2003 and similar projects), membership in the Council of Europe and overall globalization, we will probably continue to experience the trend of convergence in the future, bearing in mind the importance of quality regulation of administrative procedure as the main decision process for states' economic growth and the effectiveness of public administration activities. \\[4mm]

\cast{References}
Apelblat, M. (2011). Reflections on reforms of administrative procedures in Western Balkans. Regional School of Public Administration annual Conference: ``Modernising Administrative Procedures, meeting EU Standards'', 40--49. 

Ávila, H. (2007). \fntit{Theory of Legal Principles}. Law and Philosophy Library Volume 81. Dordrecht: Springer Netherlands: Springer. 

Barnes, J. (2010). Toward a~Third Generation of Administrative Procedure. In Rose-Ackerman, S. and Lindseth, P.L. (eds.). \fntit{Comparative Administrative Law}. cop. 2010. Cheltenham, Northampton: Edward Elgar.

Britannica Academic. (2014). \fntit{Declaration of the Rights of Man and of the Citizen}. Retrieved October 11, 2014, from  http://academic.eb.com/bps/additionalcontent/8/116846/De\-claration-of-the-Rights-of-Man-and-of-the-Citizen.

Cardona, F. (2005). Checklist for general law on administrative procedures. SIGMA.

CARDS 2003 project: Support to the Public Administration and Civil Service Reform. Retrieved April 10, 2014, from http://www.delhrv.ec.europa.eu/?lang=en\&content=287.

Davitkovski, B., and Pavlovska-Daneva, A. (2009). Realizing Citizens' Rights through the Administrative Procedure and Administrative Dispute in the Republic of Macedonia. \fntit{Hrvatska javna uprava}, 9 (1), 125--140.

Davitkovski, B., and Pavlovska-Daneva, A. (2010). The European Administrative Space as a~Challenge for Public Administration Reform in the Republic of Macedonia. \fntit{Hrvatska i~komparativna javna uprava}, 11(1), 127--146. 

\clearpage\newpage

European Commission (2001). European Commission's White Paper on European Governance, COM(2001) 428 final, 25.7.2001. Retrieved October 11, 2014, from http://aei.pitt.\\edu/1188/1/european\_governance\_wp\_COM\_2001\_428.pdf

European Commission (2012). Commission Staff Working Document, ``The Former Yugoslav Republic of Macedonia 2012 Progress Report'', Brussels 10. 10. 2012, SWD(2012) 332 final.

European Commission (1995). Preparation of the Associated Countries of Central and Eastern Europe for Integration into the Internal Market of the Union -- White Paper. COM (95) 163 final, 3 May 1995. Retrieved October 10, 2014, from http://aei.pitt.edu/1120/1/east\_en\-larg\_wp\_COM\_95\_163.pdf

Galligan, D. J., Langan, R. H. II, and Nicandrou, C. S. (1998). \fntit{Administrative Justice in the New European Democracies: Case Studies of Administrative Law and Process in Bulgaria, Estonia, Hungary, Poland and Ukraine}. Oxford: Open Society Institute, Constitutional and Legislative Policy Institute and the Centre for Socio-Legal Studies, University of Oxford.

Grafenauer, B., and Breznik, J. (2009). \fntit{Upravno pravo: procesni del [Administrative Law: procedural section]}. Ljubljana: Gospodarski vestnik.

Hirsch-Ziembiñska, M. (2007). The application of the European Code of Good Administrative Behaviour by the European Institutions. In Pursuit of Good Administration, Council of Europe, 29--30 November, Warsaw, Poland, 1--10.

Hladík, J., Kopecký, V. (2013). \fntit{Public administration reform in the Czech Republic. Research Paper Public administration reform in the Czech Republic -- March 2013}. Association for International Affairs, 2013. Retrieved October 14, 2014, from http://www.amo.cz/\\editor/image/stranky3\_soubory/rp\_par-in-ua-and-by\_2013\_2\_czech-republic.pdf

Hus, G., and Katiæ Buba¹, J. (2011). Novelties in the general administrative procedure act and its application in Croatia. Regional School of Public Administration annual Conference: ``Modernising Administrative Procedures, meeting EU Standards'', 19--29.

Iancu, D.C. (2012). Re(De)-Forming public administration: an expert outlook on reform planning in Romania. \fntit{Theoretical and Applied Economics}, XIX (8(573)), 103--124.

Ker¹evan, E. (2008). Vladavina prava in upravno pravo [Rule of Law and Administrative Law]. \fntit{Podjetje in delo} No. 6--7, 1131--1140.

Kopriæ, I. (2005). Administrative Procedures on the Territory of Former Yugoslavia. Paper presented at the Regional Workshop on Public Administration Reform and EU Integration, in Budva, Montenegro. Retrieved October 16, 2014, from http://www.sigmaweb.org/publi\-cations/36366473.pdf

Kovaè, P. (2011). Modernizing administrative procedural law in Slovenia as a~driving force of efficient and good administration. \fntit{Law and Public Management Revisited, The NISPAcee journal of public administration and policy}, 4 (2), 39--66.

Kovaè, P. (2013). Slovene Administrative reforms: at the cross-section of post-socialism, legalism and good administration. In Vintar, M.; Rosenbaum, A.; Jenei, G.; Drechsler, W. (eds.). \fntit{The Past, Present and Future of Public Administration in Central and Eastern Europe}. Bratislava: NISPAcee Press.

Künnecke, M. (2007). \fntit{Tradition and Change in Administrative Law: an Anglo-German Comparison}. Berlin, New York: Springer.

Lampe, R. (2010). \fntit{Pravo èlovekovih pravic, Sistem èlovekovih pravic v~mednarodnem, evropskem in ustavnem pravu [Human Rights Law, System of Human Rights in International, European and Constitutional Law]}. Ljubljana: Uradni list Republike Slovenije.

Loughlin, J.; Hendriks, F. and Lindström, A. (eds.) (2011). \fntit{The Oxford Handbook of Local and Regional Democracy in Europe}. New York: Oxford University Press.

Mendes, J. (2009). Good Administration in EU Law and the European Code of Good Administrative Behaviour. \fntit{European University Institute, EUI Working papers, Law 2009/09, Department of Law:} 1--19.

Nikolov, D. (2013). Public Administration and Public Administration Reform in the Western Balkan Countries. In Vintar, M.; Rosenbaum, A.; Jenei, G.; Drechsler, W. (eds.). \fntit{The Past, Present and Future of Public Administration in Central and Eastern Europe}. Bratislava: NISPAcee Press.

Novak, A. (2010). Zmuzljiva narava pravnih naèel [Elusive Nature of Legal Principles]. \fntit{Zbornik znanstvenih razprav letnik}, 70, 211--241.

Olson, J. P. (2002). The Many Faces of Europeanisation. \fntit{Journal of Common Market Studies}, 40(5), 921--952.

Pavènik, M. (1997). \fntit{Teorija prava [Theory of Law]}. Ljubljana: Cankarjeva zalo¾ba. 

Pavènik, M. (ed.) (2009). \fntit{Pravna dr¾ava. [The State Governed by the Rule of Law]}. Ljubljana: GV Zalo¾ba. 

Petrov, R., Kalinichenko, P. (2011). The europeanization of third country judiciaries through the application of the EU acquis: the cases of Russia and Ukraine. \fntit{International and Comparative Law Quarterly}, Vol. 60, 325--353.

Pollitt, C., and Bouckaert, G. (2011). \fntit{Public Management Reform, A~Comparative Analysis: New Public Management, Governance, and the Neo-Weberian State}. Oxford: Oxford University Press. 

Pouperová, O. (2014). Hauser Global Law School Program, New York University School of Law. \fntit{An Introduction to the Czech Legal System and Legal Resources Online}. Retrieved October 13, 2014, from http://www.nyulawglobal.org/globalex/czech\_republic.htm

Public Administration Reform Strategy 2010--2015, Republic of Macedonia, 2010.

Radaelli, C. M. and Featherstone, K. (eds.) (2003). \fntit{The Politics of Europeanization}. New York: Oxford University Press.

Rusch, W. (2009). Administrative Procedures in EU Member States. Paper presented at the Conference on Public Administration Reform and European Integration, SIGMA, 26--27 March, Budva, Montenegro. Retrieved October 16, 2014, from http://www.oecd.org/data\-oecd/49/34/42754772.pdf

Rusch, W. (2011). Citizens first, Good administration through general administrative procedures. Regional School of Public Administration annual Conference: ``Modernising Administrative Procedures, meeting EU Standards'', 3--18. 

Savino, M. (2010). \fntit{The Right to Open Public Administration in Europe: Emerging Legal Standards}. Sigma Paper No. 46. Paris: SIGMA, 1--40.

Schuppert, G. F. (2000). \fntit{Varwaltungswissenschaft (Verwaltung, Verwaltungsrecht, Verwaltungslehre)}. Baden-Baden: Nomos.

\clearpage\newpage

Schwarze, J. (1992). \fntit{European Administrative Law}. Luxembourg: Office for Official Publications of the European Communities; London.

Schwarze, J. (2004). Judicial Review of European Administrative Procedure. \fntit{68 Law and Contemporary Problem}, 85--105.

Schwarze, J. (2011). European Administrative Law in the Light of the Treaty of Lisbon: Introductory Remarks. Directorate General for Internal Policies, Policy Department C: Citizen's Rights and Constitutional Affairs, Legal Affairs, Workshop on EU Administrative Law: State of Play and Future Prospects (León, Spain), Briefing Notes, European Parliament, 7--27.

SIGMA. (1999). European Principles for Public Administration, Paper No. 27, CCNM/SIG\-MA/PUMA(99)44/REV1.

SIGMA. (2012). Assessment Reports on the Former Yugoslav Republic of Macedonia. Retrieved October 14, 2014, from http://www.oecd-ilibrary.org/governance/sigma-country-assessment-reports\_22260005;jsessionid=26eyiwrqh9pnm.x-oecd-live-01 

Skulová, S.; Husseini, F.; Vrbová, M; and Prokopová, K. (2011). Principles of good governance as a~factor in effective decision-making by administrative authorities and administrative courts, with accent on the principle of predictability and stability of decision-making (``Czech Republic'' case study). EGPA Annual Conference, Study Group Law and Public Administration, 7--9 September, Bucharest, Romania, 1--16. 

Sta¹a, J., and Tomá¹ek, M. (2012). Codification of administrative procedure. \fntit{The Lawyer Quarterly}, 2, 59--68.

Statskontoret (2005). \fntit{Principles of Good Administration in the Member States of the European Union}. Retrieved October 16, 2014, from http://www.statskontoret.se/upload/Publika\-tioner/2005/200504.pdf

Todevski, M. (2011). Reforming administrative procedures in the Republic of Macedonia. Regional School of Public Administration annual Conference: ``Modernising Administrative Procedures, meeting EU Standards'', 25--29.

University of Strasbourg, Centre for European Political Sociology. Hadjiisky, M. (2009). \fntit{Europeanisation and the reform of the State: The influence of the European Union in the reform of the Czech public administration (1993--2004)}. Retrieved October 14, 2014, from http://prisme.u-strasbg.fr/workingpapers/WPHadjiisky.pdf

Venice Commission (2011). Venice Commission -- European Commission for Democracy through Law. Stocktaking on the Notions of \char159 Good Governance\char158{} and \char159Good Administration\char158, Study no. 470/2008, CDL-AD(2001)009. Retrieved October 16, 2014, from http://www.venice.coe.int/webforms/documents/CDL-AD(2011)009.aspx 

Vidláková, O. (2006). \fntit{The reform and modernization of central state administration in the Czech Republic}. Retrieved October 11, 2014, from http://unpan1.un.org/intradoc/groups/\\public/documents/nispacee/unpan024319.pdf

Ziller, J. (2009, cop. 2007). The Continental System of Administrative Legality. In B. Guy, B. P. and Pierre, J. (eds.). \fntit{Handbook of Public Administration}. Los Angeles, London, New Delhi: SAGE.
